---
title: "PMBCL exome mutational analysis: Differential expression analysis of DASL data"
author: "Stacy Hung"
date: "September 22, 2018"
output: html_document
---
  
## Objectives 
  
Obtain lists of differentially expressed genes for the following comparisons:
* JAK-STAT mutated patients vs. non-JAK-STAT mutated patients
* mutated vs. non-mutated patients for each of the following genes:
    - JAK-STAT pathway: IL4R, SOCS1, STAT6, PTPN1, IL2RG, IL13RA1, JAK1, JAK2, JAK3, CISH, PTPN6, INPP5D, PTPN11
    - Top 20 MutSig genes (based on frequency): SOCS1, ITPKB, STAT6, ZNF217, GNA13, IL4R, TNFAIP3, NFKBIE, ZFP36L1, VMP1, CIITA, IRF2BP2, PTPN1, JAK1, ACTB, CD58, HIST1H1B, IRF4, DDX3X
    - IRF pathway (tentative): IRF2BP2, IRF8, IRF4, IRF1, CISH, IL13RA1, IL1RAPL2, IFNL3

NB: Expression values are per probe set, and are log-transformed and normalized (neqc). Outliers probes, probesets, and samples have been removed based on QC analysis.

## Load the data necessary for differential expression analysis

```{r}
library(limma)
library(tidyr)

# load the design of the expression data
#targets <- read.table("/Volumes/shung/projects/PMBCL_exomes/DASL/data/design-clinical+mutation_data.txt", sep = "\t", header = TRUE)
#targets <- read.table("/Volumes/shung/projects/PMBCL_exomes/DASL/data/design-jak_mutated.txt", sep = "\t", header = TRUE)
targets <- read.table("/Volumes/shung/projects/PMBCL_exomes/DASL/data/design-irf_mutated.txt", sep = "\t", header = TRUE)

# load the expression data
expr.mat <- read.table("/Volumes/shung/projects/PMBCL_exomes/DASL/analysis/DASL-expression.norm.filtered.txt", sep = "\t", header = TRUE, row.names = 1)

# extract only necessary columns
#targets <- design.data[ ,c(25:75)]
# convert design data from wide to long
targets.long <- gather(targets, gene, mutation, ACTB:ZNF217, factor_key = TRUE)

# read in patients that should be removed, and patient ids to use for exome cases
patients.to_remove <- scan(file = "/Volumes/shung/projects/PMBCL_exomes/DASL/data/patients_to_remove.txt", what = 'character')
patients.exomes <- scan(file = "/Volumes/shung/projects/PMBCL_exomes/DASL/data/patient_ids.exomes.txt", what = 'character')

# remove patients that do not have exomes
expr.exomes <- expr.mat[ , !(names(expr.mat) %in% patients.to_remove)]
# also remove extra gene columns
drops <- c("SYMBOL", "TargetID")
expr.exomes <- expr.exomes[ , !(names(expr.exomes) %in% drops)]
# rename res-ids (which start with with an "X" and have a dot instead of dash due to R constraints)
colnames(expr.exomes) <- patients.exomes

# finally, remove patient that doesn't have mutation data (to match dimensions of design matrix)
drops <- c("PA012")
expr.exomes <- expr.exomes[, !(names(expr.exomes) %in% drops)]

```

## Use limmma for differential expression analysis

```{r}
# Note to self: apparently all limma functions will accept a matrix of log2 ratios (e.g. lmFit should be able to accept a matrix)
# only look at one gene for now
targets.temp <- subset(targets.long, targets.long$gene == "ITPKB")
targets.temp$gene <- factor(targets.temp$gene)
targets.temp <- targets

#f <- paste(targets.temp$gene, targets.temp$mutation, sep = ".")
#f <- paste("mutated", targets$jak_mutated, sep = ".")
f <- paste("mutated", targets$irf_mutated, sep = ".")
f <- factor(f)
design <- model.matrix(~0+f)
colnames(design) <- levels(f)
# fit the linear model (idea: fit the systematic part of the data to distiguish it from noise)
fit <- lmFit(expr.exomes, design)
# Specify contrast matrix (what we want to compare)
#cont.matrix <- makeContrasts(mutant_V_wt="IL4R.1-IL4R.0", levels=design)
cont.matrix <- makeContrasts(mutant_V_wt="mutated.1-mutated.0", levels=design)
# extract linear model fit for the contrasts
fit2 <- contrasts.fit(fit, cont.matrix)
# eBayes represents Empirical Bayes Statistics for differential expression; it computes moderated t-statistics for each individual contrast
fit2 <- eBayes(fit2)
```

## Output data to text file - (1) top differentially expressed genes, (2) normalized expression data matrix

```{r}
#topTable(fit2, number=20)
top_table <- topTable(fit2, number = 30000)
#top_table <- topTable(fit2, number = 464, p.value = 0.05) # JAK-STAT DE genes
top_table <- topTable(fit2, number = 391, p.value = 0.05) # JAK-STAT DE genes
#write.table(top_table, "/Volumes/shung/projects/PMBCL_exomes/DASL/analysis/top_table-jak_stat-mutant_V_wt.txt", sep="\t", quote = FALSE)
write.table(top_table, "/Volumes/shung/projects/PMBCL_exomes/DASL/analysis/top_table-irf-mutant_V_wt.txt", sep="\t", quote = FALSE)

# optional: print out expression values
selected <- p.adjust(fit2$p.value, method = "BH") < 0.05
#esetSel <- expr.exomes[selected,]
expression_values <- expr.exomes[selected,]
#expression_values <- exprs(esetSel)
#write.table(expression_values, "/Volumes/shung/projects/PMBCL_exomes/DASL/analysis/top_table-jak_stat-mutated_V_wt-expression_values.txt", sep="\t", quote = FALSE)
write.table(expression_values, "/Volumes/shung/projects/PMBCL_exomes/DASL/analysis/top_table-irf-mutated_V_wt-expression_values.txt", sep="\t", quote = FALSE)
```

## Visualize the differentially expressed genes

```{r}
library(reshape)
library(ggplot2)
library(ggrepel)
library(RColorBrewer)
library(pheatmap)

# first add gene names to the data
source ("/Volumes/shung/scripts/addNewData.R")
allowedVars <- c("gene")
top_table$probeset_id <- row.names(top_table)
top_table <- addNewData("/Volumes/shung/projects/PMBCL_exomes/DASL/data/dataNew-probeset_to_gene.csv", 
                        top_table, 
                        allowedVars)

# order the resutlts by adjusted p-value
results_de.ordered <- top_table[order(top_table$adj.P.Val), ]
top20.de_genes <- results_de.ordered[1:20, ]$gene
top20.de_probesets <- row.names(results_de.ordered[1:20, ])

# now we can extract the normalized expression values for these top 20 genes:
top20.sigDE.norm_expr <- data.frame(expression_values[top20.de_probesets, ])
# create a column with the gene names (from row names)
top20.sigDE.norm_expr$probeset_id <- row.names(top20.sigDE.norm_expr)
top20.sigDE.norm_expr <- addNewData("/Volumes/shung/projects/PMBCL_exomes/DASL/data/dataNew-probeset_to_gene.csv",
                                    top20.sigDE.norm_expr, allowedVars)
# melt data to use for ggplotting
top20.sigDE.norm_expr.melted <- melt(top20.sigDE.norm_expr)
colnames(top20.sigDE.norm_expr.melted) <- c("probeset_id", "gene", "patient_id", "normalized_log2_expr")
# map mutation status for patient_id so that it can be used for plotting
allowedVars <- c("mutated_patient")
top20.sigDE.norm_expr.melted <- addNewData("/Volumes/shung/projects/PMBCL_exomes/DASL/analysis/jak-stat_pathway/dataNew-jakstat_pathway-patient_id_to_mutation_status.csv", top20.sigDE.norm_expr.melted, allowedVars)

# plot expression for each gene - differentiate between mutated and non-mutated patients
ggplot(top20.sigDE.norm_expr.melted) +
  geom_point(aes(x=gene, y=normalized_log2_expr, color = mutated_patient), position=position_jitter(w=0.1,h=0)) +
  scale_y_log10()+
  xlab("Genes") +
  ylab("Normalized log2 expression") +
  ggtitle("Top 20 significant DE genes") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  theme(plot.title = element_text(hjust = 0.5))

```

## Volcano plot of the differentially expressed genes

```{r}
# obtain logical vector containing information on whether p-adjusted values are < 0.05
threshold_DE.logFC <- abs(top_table$logFC) > 0.75
threshold_DE.adjPval <- abs(top_table$adj.P.Val) < 0.05
# add logical vector as a column to results table
top_table$threshold_DE.adjPval <- threshold_DE.adjPval
top_table$threshold_logFC <- threshold_DE.logFC

# label genes that have an adjusted p-value < 0.05, AND absolute logFC > 0.75
results_de.ordered_logFC <- top_table[rev(order(abs(top_table$logFC))), ]
# then sort by TRUE threshold_logFC, and TRUE threshold_DE.adjPval
results_de.ordered_logFC_and_adjPval <- results_de.ordered_logFC[
  with(results_de.ordered_logFC, rev(order(threshold_DE.adjPval, threshold_logFC))),
]

# find out how many genes meet the conditions specified above
num_sig.genes <- nrow(subset(results_de.ordered_logFC_and_adjPval , 
                    threshold_DE.adjPval == TRUE & threshold_logFC == TRUE))

# give labels to those significant genes
results_de.ordered_logFC_and_adjPval$genelabels <- ""
results_de.ordered_logFC_and_adjPval$genelabels[1:num_sig.genes] <- results_de.ordered_logFC_and_adjPval$gene[1:num_sig.genes]

# volcano plot
y_axis_label <- expression(paste("-",Log[10], "(P-value)"))
#x_axis_label <- expression(paste(Log[2], " fold change (JAK-STAT WT - Mutant)"))
x_axis_label <- expression(paste(Log[2], " fold change (IRF WT - Mutant)"))
#plot_title <- "Effect of mutations in Jak-Stat Pathway on gene expression"
plot_title <- "Effect of mutations in IRF Pathway on gene expression"

# volcano plot with gene names
p2 <- ggplot(results_de.ordered_logFC_and_adjPval) +
  geom_point(aes(x=logFC, y=-log10(adj.P.Val), colour=threshold_DE.adjPval)) +
  geom_text_repel(aes(x = logFC, y = -log10(adj.P.Val), label = results_de.ordered_logFC_and_adjPval$genelabels)) +
  ggtitle(plot_title) +
  xlab(x_axis_label) + 
  ylab(y_axis_label) +
  theme(legend.position = "none",
        plot.title = element_text(size = rel(1.5), hjust = 0.5),
        axis.title = element_text(size = rel(1.25))) +
  geom_hline(yintercept = -log10(0.05), linetype="dashed") +
  scale_color_manual(breaks = c("FALSE", "TRUE"),
                        values=c("gray", "darkblue"))

```

## MA plot
```{r}
library(DESeq2)

ma <- results_de.ordered_logFC_and_adjPval[, c("AveExpr", "logFC", "threshold_logFC")]
DESeq2::plotMA(ma, ylim=c(-2,2))
```

## Heatmap of significantly differentially expressed genes

```{r}
# extract data for significant genes
sigDE <- subset(results_de.ordered_logFC_and_adjPval, results_de.ordered_logFC_and_adjPval$adj.P.Val < 0.05)

# extract expression data for signficant genes
expr.sigDE <- expr.exomes[row.names(sigDE),]

# draw heatmap using pheatmap
annotation <- data.frame()

# set colour palette
heat_colours <- brewer.pal(4, "RdGy")

# generate heatmap
p <- ggplot(expr.sigDE, aes(variable, Name)) + geom_tile(aes(fill = rescale), colour = "white") + scale_fill_gradient(low = "white", high = "steelblue")
#pheatmap(expr.sigDE, color = heat_colours, cluster_rows = TRUE, show_rownames = FALSE, scale = "row")

```
