---
title: "SNV Heatmap"
author: "Stacy Hung"
date: "February 19, 2018"
output: html_document
---

This script generates a figure, similar to Fig. 1, but with recurrently mutated genes (n = 73).

Version 2: The most recurrently mutated gene, HRNR, with frequency higher than SOCS, was found to be an outlier and was removed from the list of recurrently mutated genes (thus n=72).

## Installation of GenVisR (optional)
```{r}
# NB: if you need to install this package, install it using devtools from the github source!!!
# The bioconductor version of GenVisR is buggy (not compatible with required libraries)
# Installing GenVisR may also require the installation of the VariantAnnotation pacakge (from bioconductor)
library(VariantAnnotation)
library(GenVisR)
```

## Read in the necessary data

```{r}
# read in mutation data
mutations <- read.table("/Volumes/shung/projects/PMBCL_exomes/Mutsig/mutsig_input/mutation_file/singles_and_pairs/singles_and_pairs.snvs_and_indels.somatic_modified_effects.removed_non_genes.singles_AF_filtered.maf", sep = "\t", header = TRUE, fill = TRUE)

# load top recurrently mutated genes
supptable2 <- read.table("/Volumes/shung/projects/PMBCL_exomes/manuscript/current/supplementary/supp_table_2.txt", sep = "\t", header = TRUE, fill = TRUE)
recurrent.genes <- as.vector(unique(supptable2$gene_hugo))

# load any metadata associated with the dataset
gender <- read.table("/Volumes/shung/projects/PMBCL_exomes/data/gender.txt", sep = "\t", header = TRUE)
lymph3cx <- read.table("/Volumes/shung/projects/PMBCL_exomes/lymphc3x/lymph3cx.txt", sep = "\t", header = TRUE)
mediastinal <- read.table("/Volumes/shung/projects/PMBCL_exomes/data/mediastinal.txt", sep = "\t", header = TRUE)
#ninep <- read.table("/Volumes/shung/projects/PMBCL_exomes/data/9p.txt", sep = "\t", header = TRUE)
```

## Oncoplots

```{r}
library(maftools)
library(gridExtra)

#####################################################
# Split patients into discovery and extension cohorts
#####################################################

# standard list of significantly mutated genes, along with clinical data annotations 
clinData.df <- read.table("/Volumes/shung/projects/PMBCL_exomes/data/clinData.v4.txt", sep = "\t", header = TRUE)

# ALL mutations (not just ones in significant genes)
mutations.maf <- read.maf("/Volumes/shung/projects/PMBCL_exomes/snv_indel_analysis/maf/mutation_summary-paired_and_unpaired.maf", clinicalData = clinData.df)

# load top recurrently mutated genes
supptable2 <- read.table("/Volumes/shung/projects/PMBCL_exomes/manuscript/current/supplementary/supp_table_2.txt", sep = "\t", header = TRUE, fill = TRUE)
recurrent.genes <- as.vector(unique(supptable2$gene))

# colours for annotation tracks
annot.colors <- list(
  Cohort = c("Discovery"="red", "Extension"="grey"),
  Gender = c("Female"="palevioletred1", "Male"="royalblue1"),
  Lymph3Cx = c("PMBL"="mediumseagreen", 
                            "ABC"="salmon1", 
                            "GCB"="cyan3", 
                            "Unclear"="black", 
                            "No_data"="white"),
  Biopsy_Site = c("Mediastinal"="mediumpurple1", "Non-mediastinal"="black"))

#patients.ordered <- c("PA002", "PA014", "PA070", "PA011", "PA010", "PA072", "PA056", "PA045", "PA016", "PA018", "PA009", "PA069", "PA003", "PA073", "PA021", "PA017", "PA013", "PA001", "PA015", "PA034", "PA031", "PA054", "PA076", "PA097", "PA042", "PA035", "PA067", "PA082", "PA028", "PA075", "PA049", "PA077", "PA038", "PA022", "PA062", "PA089", "PA093", "PA053", "PA052", "PA005", "PA064", "PA027", "PA037", "PA063", "PA096", "PA065", "PA026", "PA086", "PA058", "PA068", "PA006", "PA079", "PA024", "PA088", "PA020", "PA033", "PA008", "PA004", "PA080", "PA083", "PA060", "PA057", "PA092", "PA055", "PA044", "PA023", "PA059", "PA036", "PA039", "PA048", "PA007", "PA061", "PA041", "PA043", "PA047", "PA040", "PA029", "PA025", "PA030", "PA019", "PA085", "PA071", "PA095", "PA074", "PA078", "PA084", "PA081", "PA032", "PA066", "PA091", "PA050", "PA094", "PA051", "PA087")

# without HRNR
patients.ordered <- c("PA070", "PA011", "PA010", "PA072", "PA056", "PA045", "PA016", "PA002", "PA018", "PA009", "PA069", "PA014", "PA003", "PA073", "PA021", "PA017", "PA013", "PA001", "PA015", "PA034", "PA032", "PA031", "PA054", "PA076", "PA097", "PA042", "PA035", "PA067", "PA066", "PA082", "PA028", "PA075", "PA049", "PA091", "PA077", "PA038", "PA022", "PA062", "PA089", "PA093", "PA053", "PA050", "PA052", "PA005", "PA064", "PA027", "PA037", "PA063", "PA096", "PA094", "PA065", "PA026", "PA086", "PA058", "PA068", "PA006", "PA079", "PA024", "PA088", "PA020", "PA033", "PA008", "PA004", "PA051", "PA080", "PA083", "PA060", "PA057", "PA092", "PA055", "PA044", "PA023", "PA059", "PA036", "PA039", "PA048", "PA007", "PA061", "PA041", "PA043", "PA047", "PA040", "PA029", "PA025", "PA030", "PA019", "PA085", "PA071", "PA095", "PA074", "PA078", "PA084", "PA087", "PA081")

#genes.to_ignore <- c("MUC12")

oncoplot(maf = mutations.maf, 
         genes = recurrent.genes, 
         clinicalFeatures = c("Cohort", "Gender", "Lymph3Cx", "Biopsy_Site"),
         annotationColor = annot.colors, 
         legendFontSize = 12, showTumorSampleBarcodes = TRUE,
         drawColBar = FALSE, sampleOrder = patients.ordered, removeNonMutated = FALSE)

```

## Generate matrix to manually define order of patients in oncoplot using excel to sort

```{r}
library(dplyr)

mutations <- read.table("/Volumes/shung/projects/PMBCL_exomes/snv_indel_analysis/maf/mutation_summary-paired_and_unpaired.maf", sep = "\t", header = TRUE, fill = TRUE)

# only look at whether a patient is mutated or not (hence unique)
mutations <- unique(mutations[, c("Hugo_Symbol", "Tumor_Sample_Barcode")])

# load top recurrently mutated genes
supptable2 <- read.table("/Volumes/shung/projects/PMBCL_exomes/manuscript/current/supplementary/supp_table_2.txt", sep = "\t", header = TRUE, fill = TRUE)
recurrent.genes <- as.vector(unique(supptable2$gene))

mutations <- filter(mutations, mutations$Hugo_Symbol %in% recurrent.genes)
mutations$Hugo_Symbol <- factor(mutations$Hugo_Symbol)

mutation.matrix <- table(mutations$Hugo_Symbol, mutations$Tumor_Sample_Barcode)

write.table(mutation.matrix, "/Volumes/shung/projects/PMBCL_exomes/snv_indel_analysis/mutation_matrix.0_or_1.recurrent_genes.txt", sep = "\t", quote = FALSE)

```
