---
title: "mutation_frequency"
author: "Stacy Hung"
date: "January 10, 2017"
output: html_document
---

This script generates Fig4, which contrasts mutational frequencies between PMBCL, DLBCL, and HL.

Version 2 figures have been generated for panels B and C, which sort frequencies by average DLBCL and HL, respectively.

## Load required libraries

```{r}
library(ggplot2)
library(gridExtra)
library(grid)
library(lattice)
```

## A) Mutation frequencies of PMBCL MutSig genes compared to DLBCL (mutations obtained from strelka on Reddy dataset)

```{r}
# for significant genes in PMBCL:
pmbcl.freq <- read.table("/Volumes/shung/projects/PMBCL_exomes/frequency_analysis/pmbcl-sig.gene_counts.txt", sep = "\t", header = TRUE)
dlbcl.freq <- read.table("/Volumes/shung/projects/PMBCL_exomes/frequency_analysis/dlbcl_shipp_reddy/reddy-pmbcl_sig_genes-gene_counts.txt", sep = "\t", header = TRUE)

NUM_EXOMES_PMBCL = 94
NUM_EXOMES_DLBCL = 985

# calculate proportion of cases that are mutated
pmbcl.freq$proportion <- pmbcl.freq$Freq / NUM_EXOMES_PMBCL * 100
dlbcl.freq$proportion <- dlbcl.freq$Freq / NUM_EXOMES_DLBCL * 100

# label the datasets
pmbcl.freq$dataset <- "Our study (n=94)"
dlbcl.freq$dataset <- "Reddy et al. (n=985)"

# Get additional DLBCL (non-driver genes) from /projects/mdata_dart/Analysis/Exome/StrelkaPipeline/
# Frequency plot: our dataset (pairs+singles) and the Reddy et al. dataset re-analyzed with Strelka by the Morin Lab
freq.data.pmbcl_V_dlbcl <- rbind(pmbcl.freq, dlbcl.freq)
colnames(freq.data.pmbcl_V_dlbcl) <- c("gene", "freq", "proportion", "dataset")
  
# NEW: remove genes that are mutated < 5% (SPINK1, KLF9, LHFPL3)
genes.to_exclude <- c("SPINK1", "KLF9", "LHFPL3")
freq.data.pmbcl_V_dlbcl <- subset(freq.data.pmbcl_V_dlbcl, 
                                     !(freq.data.pmbcl_V_dlbcl$gene %in% genes.to_exclude))

freq.data.pmbcl <- subset(freq.data.pmbcl_V_dlbcl, freq.data.pmbcl_V_dlbcl$dataset == "Our study (n=94)")

# reorder genes by increasing frequency in PMBCL dataset
freq.data.pmbcl_V_dlbcl$gene <- factor(freq.data.pmbcl_V_dlbcl$gene, 
                                       levels = freq.data.pmbcl_V_dlbcl$gene[order(freq.data.pmbcl$proportion)])

panelA.pmbcl_V_dlbcl <- ggplot(data=freq.data.pmbcl_V_dlbcl, aes(x = gene, y = proportion, fill=dataset)) +
  geom_bar(stat="identity", position=position_dodge(), width = 0.8) +
  ylab("Cases mutated (%)") + xlab("") +
  theme_bw() +
  theme(legend.position="bottom") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, size=14, vjust=0.5, face = "italic"),
        axis.text.y  = element_text(size=14), 
        axis.title.x = element_text(size = 18), 
        axis.title.y = element_text(size = 18), 
        legend.text = element_text(size=14), 
        plot.margin = unit(c(1,1,1,1), "cm")) +
  scale_fill_manual(values = c("steelblue4","firebrick1"),
                    breaks = c("Our study (n=94)", 
                               "Reddy et al. (n=985)"),
                    labels = c("Our study (n=94)",
                               "Reddy et al. (n=985)"),
                    name = "")

## customized version for CCRC2019 rapid fire presentation
ccrc.pmbcl_V_dlbcl <- ggplot(data=freq.data.pmbcl_V_dlbcl, aes(x = gene, y = proportion, fill=dataset)) +
  geom_bar(stat="identity", position=position_dodge(), width = 0.8) +
  ylab("") + xlab("") +
  theme_bw() +
  theme(legend.position="bottom") +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
  theme(axis.title.x=element_blank(), axis.text.x=element_blank(), axis.ticks.x=element_blank()) + 
  theme(axis.text.y  = element_text(size=14), 
        axis.title.y = element_text(size = 18), 
        legend.text = element_text(size=14), 
        plot.margin = unit(c(1,1,1,1), "cm")) +
  scale_fill_manual(values = c("steelblue4","darkolivegreen3"),
                    name = "") +
  theme(legend.position = "none")

```

## B) Frequency plot comparing our dataset (all samples) to two DLBCL datasets (Reddy and Shipp)

```{r}
library(tidyr)

# read in frequency data for the 3 datasets
freq.data <- read.table("/Volumes/shung/projects/PMBCL_exomes/frequency_analysis/dlbcl_shipp_reddy/PMBCL_DLBCL_Shipp_Reddy-gene_proportions_complete-Shipp+Reddy_genes.txt", sep = "\t", header = TRUE)

# order bars by specified dataset:
freq.data$dataset <- factor(freq.data$dataset, levels = c("Our study (n=94)", "Reddy et al. (n=1001)", "Chapuy, Shipp (n=304)"))

# version 1: reorder by value of PMBCL frequency
#freq.data.pmbcl <- subset(freq.data, freq.data$dataset == "Our study (n=94)")


# version 2: reorder by average DLBCL frequency
# convert to wide so we can sort by the pathology of interest
freq.data.wide <- spread(freq.data, dataset, proportion)
freq.data.wide$mean_dlbcl <- rowMeans(freq.data.wide[c("Reddy et al. (n=1001)", "Chapuy, Shipp (n=304)")])
freq.data.wide <- freq.data.wide[order(freq.data.wide$mean), ]
freq.data <- gather(freq.data.wide, dataset, proportion, `Our study (n=94)`:`Chapuy, Shipp (n=304)`)
freq.data$gene <- factor(freq.data$gene, levels = freq.data$gene[order(freq.data.wide$mean_dlbcl)])

panelB.dlbcl_V_dlbcl <- ggplot(data=freq.data, aes(x = gene, y = proportion, fill=dataset)) +
  geom_bar(stat="identity", position=position_dodge(), width = 0.8) +
  ylab("Cases mutated (%)") + xlab("") +
  theme_bw() +
  theme(legend.position="bottom") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, size=14, vjust=0.5, face = "italic"),
        axis.text.y  = element_text(size=14), 
        axis.title.x = element_text(size = 18), 
        axis.title.y = element_text(size = 18), 
        legend.text = element_text(size=14), 
        plot.margin = unit(c(1,1,1,1), "cm")) +
  scale_fill_manual(values = c("orange", "steelblue4", "firebrick1"),
                    breaks = c("Our study (n=94)", 
                               "Reddy et al. (n=1001)",
                               "Chapuy, Shipp (n=304)"),
                    labels = c("Our study (n=94)",
                               "Reddy et al. (n=1001)",
                               "Chapuy, Shipp (n=304)"),
                    name = "")

# for version 1, use this ordering for fill colours: 
# scale_fill_manual(values = c("steelblue4", "firebrick1", "orange"),

## customized version for CCRC2019
freq.data$gene <- factor(freq.data$gene, levels = freq.data$gene[order(freq.data.wide$`Our study (n=94)`)])

ccrc.dlbcl_V_dlbcl <- ggplot(data=freq.data, aes(x = gene, y = proportion, fill=dataset)) +
  geom_bar(stat="identity", position=position_dodge(), width = 0.8) +
  ylab("Cases mutated (%)") + xlab("") +
  theme_bw() +
  theme(legend.position="bottom") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, size=14, vjust=0.5, face = "italic"),
        axis.text.y  = element_text(size=14), 
        axis.title.x = element_text(size = 18), 
        axis.title.y = element_text(size = 18), 
        legend.text = element_text(size=14), 
        plot.margin = unit(c(1,1,1,1), "cm")) +
  scale_fill_manual(values = c("darkseagreen3", "steelblue4", "darkseagreen2"),
                    breaks = c("Our study (n=94)", 
                               "Reddy et al. (n=1001)",
                               "Chapuy, Shipp (n=304)"),
                    labels = c("Our study (n=94)",
                               "Reddy et al. (n=1001)",
                               "Chapuy, Shipp (n=304)"),
                    name = "") +
  theme(axis.title.x=element_blank(), axis.text.x=element_blank(), axis.ticks.x=element_blank()) + 
  theme(legend.position = "none")

```


## C) Frequency plot comparing our dataset (all samples) to the HL Tiacci et al. dataset

```{r}
freq.data.hl <- read.table("/Volumes/shung/projects/PMBCL_exomes/frequency_analysis/hl_tiacci/hl_vs_pmbcl-tiacci_sig_genes.gene_proportions.txt", sep = "\t", header = TRUE)

# version 1: keep order specified in input data
#freq.data.hl$gene <- factor(freq.data.hl$gene, levels = freq.data.hl$gene)

# version 2: reorder genes by increasing frequency in HL dataset
# convert to wide so we can sort by the pathology of interest
freq.data.hl.wide <- spread(freq.data.hl, dataset, proportion)
freq.data.hl.wide <- freq.data.hl.wide[order(freq.data.hl.wide$`Tiacci et al. (n=34)`, freq.data.hl.wide$`Our study (n=94)`), ]
freq.data.hl <- gather(freq.data.hl.wide, dataset, proportion, `Our study (n=94)`:`Tiacci et al. (n=34)`)
freq.data.hl$gene <- factor(freq.data.hl$gene, levels = freq.data.hl$gene[order(freq.data.hl.wide$`Tiacci et al. (n=34)`)])


#freq.data.hl$gene <- factor(freq.data.hl$gene, levels = freq.data.hl$gene[order(pmbcl.freq$proportion)])

panelC.hl_V_pmbcl <- ggplot(data=freq.data.hl, aes(x = gene, y = proportion, fill=dataset)) +
  geom_bar(stat="identity", position=position_dodge(), width = 0.8) +
  ylab("Cases mutated (%)") + xlab("") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, size=14, vjust=0.5, face = "italic"),
        axis.text.y  = element_text(size=14), 
        axis.title.x = element_text(size = 18), 
        axis.title.y = element_text(size = 18), 
        legend.text = element_text(size=14), 
        plot.margin = unit(c(1,1,1,1), "cm")) +
  scale_fill_manual(values = c("steelblue4","limegreen"),
                    breaks = c("Our study (n=94)", 
                               "Tiacci et al. (n=34)"),
                    labels = c("Our study (n=94)",
                               "Tiacci et al. (n=34)"),
                    name = "") +
  theme(legend.position="bottom")

```


## Arrange all the plots together in one plot

```{r}
grid.arrange(panelA.pmbcl_V_dlbcl, panelB.dlbcl_V_dlbcl, panelC.hl_V_pmbcl, ncol=1)
```
