---
title: "SNV Heatmap"
author: "Stacy Hung"
date: "February 19, 2018"
output: html_document
---

This script is used to generate plots for visualizing mutation data for the PMBCL exomes dataset.

## Installation of GenVisR (optional)
```{r}
# NB: if you need to install this package, install it using devtools from the github source!!!
# The bioconductor version of GenVisR is buggy (not compatible with required libraries)
# Installing GenVisR may also require the installation of the VariantAnnotation pacakge (from bioconductor)
source("https://bioconductor.org/biocLite.R")
biocLite("VariantAnnotation")
library("VariantAnnotation")

install.packages("devtools")
library(devtools)
install_github("griffithlab/GenVisR")
library("GenVisR")
```

## Read in the necessary data

```{r}
# read in mutation data
mutations <- read.table("/Volumes/shung/projects/PMBCL_exomes/Mutsig/mutsig_input/mutation_file/singles_and_pairs/singles_and_pairs.snvs_and_indels.somatic_modified_effects.removed_non_genes.singles_AF_filtered.maf", sep = "\t", header = TRUE, fill = TRUE)

#mutations <- read.table("/Volumes/shung/projects/PMBCL_exomes/Mutsig/mutsig_input/mutation_file/maftools/paired_and_unpaired.snvs_and_indels.sig_genes_v3.hugo_gene.AF_filtered.gene_filtered.maf", sep = "\t", header = TRUE)

# load mutation data (MAF - filtering done on VarScan+MuTect+Strelka *plus* silent mutations) - excludes UTR mutations
#mutations.maf <- read.table("/Volumes/shung/projects/PMBCL_exomes/Mutsig/mutsig_input/mutation_file/pairs.snvs_and_indels.somatic_modified_effects.maf", sep = "\t", header = TRUE)

# load MutSig output (list of significantly mutated genes)
sig.genes <- read.table("/Volumes/shung/projects/PMBCL_exomes/Mutsig/mutsig_output/pairs/ensembl_ids_with_coverage_data_v3/pairs.snvs_indels.ensembl.coverage.v3.sig_genes.refseq.txt", sep = "\t", header = TRUE, na.strings = "NA")

# load any metadata associated with the dataset
gender <- read.table("/Volumes/shung/projects/PMBCL_exomes/data/gender.txt", sep = "\t", header = TRUE)
lymph3cx <- read.table("/Volumes/shung/projects/PMBCL_exomes/data/lymph3cx.txt", sep = "\t", header = TRUE)
ninep <- read.table("/Volumes/shung/projects/PMBCL_exomes/data/9p.txt", sep = "\t", header = TRUE)
```

## Now subset the data for appropriate visualization

```{r}
# get significant genes
sig.genes.list <- subset(sig.genes, sig.genes$p < 0.05)

# filter out silent mutations
library(dplyr)    # filter
mutations <- filter(mutations, !Variant_Classification == 'Silent')
mutations$Variant_Classification <- factor(mutations$Variant_Classification)

# first merge the two datasets
keeps <- c("Hugo_Symbol", "Tumor_Sample_Barcode", "Variant_Classification", "Chromosome", "Start_Position", "Reference_Allele", "Tumor_Seq_Allele1", "Tumor_Seq_Allele2")
mutations <- mutations[keeps]

# only focus on significant genes
mutations.sig <- mutations[mutations$Hugo_Symbol %in% sig.genes.list$gene_hugo, ]
mutations.sig$gene <- factor(mutations.sig$Hugo_Symbol)
rm(mutations)

# calculate the End_Position column for maftools maf format (End_Position = length(Reference) - start + 1)
mutations.sig$End_Position <- nchar(as.character(mutations.sig$Reference_Allele)) + as.numeric(mutations.sig$Start_Position) - 1

# map Variant_Classification to a Variant_Type field (another column we need for maftools)
# Variant_Type = {INS, DEL, SNP, DNP, TNP, ONP}
class.to.type <- c("Missense_Mutation"="SNP",
                   "Nonsense_Mutation"="SNP",
                   "Splice_Site"="SNP",
                   "Nonstop_Mutation"="SNP",
                   "Targeted_Region"="SNP",
                   "Frame_Shift_Del"="DEL",
                   "Frame_Shift_Ins"="INS"
                   )
mutations.sig$Variant_Type <- revalue(mutations.sig$Variant_Classification, class.to.type)

# write out as plottable maf
write.table(mutations.sig, "/Volumes/shung/projects/PMBCL_exomes/Mutsig/mutsig_input/mutation_file/maftools/paired_and_unpaired.snvs_and_indels.sig_genes_v3.refseq.singles_AF_filtered.maf", sep = "\t", row.names = FALSE, quote = FALSE)

```

### Generate AA change column from HGVS annotations to use for lollipop plots

```{r}
library(tidyr)    # separate
library(DataCombine) # find and replace

mutations <- read.table("/Volumes/shung/projects/PMBCL_exomes/Mutsig/mutsig_input/mutation_file/maftools/paired_and_unpaired.snvs_and_indels.sig_genes_v3.hugo_gene.AF_filtered.gene_filtered+IRF1_AA_Change.txt", sep = "\t", header = TRUE, fill = TRUE)

mutations <- read.table("/Volumes/shung/projects/PMBCL_exomes/Mutsig/mutsig_input/mutation_file/maftools/paired_and_unpaired.snvs_and_indels.sig_genes.AA_change.txt", sep = "\t", header = TRUE, fill = TRUE)

mutations <- read.table("/Volumes/shung/projects/PMBCL_exomes/Mutsig/mutsig_input/mutation_file/maftools/paired_and_unpaired.snvs_and_indels.sig_genes_v3.hugo_gene.AF_filtered.gene_filtered.with_AA_Change.txt", sep = "\t", header = TRUE)

mutations <-
  separate(data = mutations,
           col = HGVS,
           into = c("Protein_Change", "CDS_Change"),
           sep = "/",
           remove = TRUE,
           fill = "left")

# mappings for 3-letter amino acids to 1-letter amino acid codes
AA.replaces <- data.frame(from = c("Ala", "Arg", "Asn", "Asp", "Asx", "Cys", "Glu", "Gln", "Glx", "Gly", "His", 
                                   "Ile", "Leu", "Lys", "Met", "Phe", "Pro", "Ser", "Thr", "Trp", "Tyr", "Val"), 
                          to = c("A", "R", "N", "D", "B", "C", "E", "Q", "Z", "G", "H", 
                                 "I", "L", "K", "M", "F", "P", "S", "T", "W", "Y", "V"))

mutations <- FindReplace(data = mutations, 
                                    Var = "Protein_Change",
                                    replaceData = AA.replaces,
                                    from = "from",
                                    to = "to",
                                    exact = FALSE)

write.table(mutations, "/Volumes/shung/projects/PMBCL_exomes/Mutsig/mutsig_input/mutation_file/maftools/paired_and_unpaired.snvs_and_indels.sig_genes_v3.hugo_gene.AF_filtered.gene_filtered+IRF1_AA_Change.maf", sep = "\t", quote = FALSE, row.names = FALSE)

write.table(mutations, "/Volumes/shung/projects/PMBCL_exomes/Mutsig/mutsig_input/mutation_file/maftools/paired_and_unpaired.snvs_and_indels.sig_genes.AA_change.maf", sep = "\t", quote = FALSE, row.names = FALSE)

write.table(mutations, "/Volumes/shung/projects/PMBCL_exomes/Mutsig/mutsig_input/mutation_file/maftools/paired_and_unpaired.snvs_and_indels.sig_genes_v3.hugo_gene.AF_filtered.gene_filtered.with_AA_Change.maf", sep = "\t", quote = FALSE, row.names = FALSE)

```

## Oncoplots

```{r}
source("https://bioconductor.org/biocLite.R")
biocLite("maftools")
library(maftools)
library(gridExtra)

# standard list of significantly mutated genes, along with clinical data annotations
#clinData.df <- read.table("/Volumes/shung/projects/PMBCL_exomes/data/clinData.txt", sep = "\t", header = TRUE)
#mutations.maf <- read.maf("/Volumes/shung/projects/PMBCL_exomes/Mutsig/mutsig_input/mutation_file/maftools/paired_and_unpaired.snvs_and_indels.sig_genes_v3.hugo_gene.AF_filtered.gene_filtered.with_VAF.unique.maf", clinicalData = clinData.df)

# ALL mutations (not just ones in significant genes)
#mutations.maf <- read.maf("/Volumes/shung/projects/PMBCL_exomes/snv_indel_analysis/maf/mutation_summary-paired_and_unpaired.maf", clinicalData = clinData.df)

# ALL mutations *including* UTR mutations (and all genes)
mutations.maf <- read.maf("/Volumes/shung/projects/PMBCL_exomes/snv_indel_analysis/maf/mutation_summary-paired_and_unpaired_with_UTRs.maf", clinicalData = clinData.df, vc_nonSyn = c("Frame_Shift_Del", "Frame_Shift_Ins", "Splice_Site", "Translation_Start_Site","Nonsense_Mutation", "Nonstop_Mutation", "In_Frame_Del","In_Frame_Ins", "Missense_Mutation", "3'UTR", "5'UTR"))

# significant mutations plus IRF (use this mutation list for highlighting IRF pathway genes or overlaying CN)
#mutations.maf <- read.maf("/Volumes/shung/projects/PMBCL_exomes/Mutsig/mutsig_input/mutation_file/maftools/paired_and_unpaired.snvs_and_indels.sig_genes_v3.hugo_gene.AF_filtered.gene_filtered+IRF1.maf", 
#                          clinicalData = clinData.df,
#                          gisticAllLesionsFile = "/Volumes/shung/projects/PMBCL_exomes/gistic/output/drop_low_coverage-filtered_949-arm_peel-threshold_0.3-fdr_0.05-max_seg_1000-FINAL_FINAL_FINAL/gistic_segments.all_lesions.conf_90.txt", 
#                          gisticAmpGenesFile = "/Volumes/shung/projects/PMBCL_exomes/gistic/output/drop_low_coverage-filtered_949-arm_peel-threshold_0.3-fdr_0.05-max_seg_1000-FINAL_FINAL_FINAL/gistic_segments.amp_genes.conf_90.txt", 
#                          gisticDelGenesFile = "/Volumes/shung/projects/PMBCL_exomes/gistic/output/drop_low_coverage-filtered_949-arm_peel-threshold_0.3-fdr_0.05-max_seg_1000-FINAL_FINAL_FINAL/gistic_segments.del_genes.conf_90.txt")

sig.genes <- c("SOCS1", "ITPKB", "STAT6", "GNA13", "NFKBIE", "PTPN1", "ZNF217", "IL4R", "ACTB", "HIST1H3D", "CD58", "IRF8", "SGPP1", "TNFAIP3", "GCSAM", "JUNB", "ZFP36L1", "RGS1", "HIST1H2BC", "HIST1H1C", "JAK1", "MS4A1", "SRSF1", "HIST1H1B", "EZH2", "IFNL3", "GDF11", "KLF9", "IL1RAPL2", "SERPINB3", "IGLJ2", "VMP1", "MYC", "LHFPL3", "TSPAN19", "IRF2BP2", "DDX3X", "CIITA", "IRF4", "CXCR5", "EPHA7", "IL13RA1", "CISH", "SPINK1", "TRAF3", "HIST1H4E", "HIST1H4J", "MAGEC2", "IRX2", "FPGT")

# only with copy number:
sig.genes <- c("SOCS1", "HIST1H1C", "TNFAIP3", "EZH2")

# colours for annotation tracks
annot.colors <- list(Gender = c("Female"="palevioletred1", "Male"="royalblue1"),
               Lymph3Cx = c("PMBL"="mediumseagreen", 
                            "ABC"="salmon1", 
                            "GCB"="cyan3", 
                            "Unclear"="black", 
                            "No_data"="white"),
               Status_9p = c("amplification"="firebrick3", "gain"="firebrick1", "neutral"="black"))
#R.CHOP = c("Yes"="mediumpurple1", "No"="khaki2"))

# Standard mutation waterfall plot (oncoplot style), and with clinical annotations
oncoplot(maf = mutations.maf, genes = sig.genes, 
         clinicalFeatures = c("Gender", "Lymph3Cx", "Status_9p"),
         annotationColor = annot.colors, 
         legendFontSize = 12, drawColBar = FALSE)

# oncoplot for IRF pathway

#oncoplot(maf = mutations.maf, genes = c("IRF2BP2", "IRF8", "IRF4", "IRF1", "CISH", "IL13RA1", "IL1RAPL2", "IFNL3"))

oncoplot(maf = mutations.maf, genes = c("TLR2", "TLR4", "TLR5", "MYD88", "RELA", "NFKB1", "TRAM1", "TICAM1", "TRAF6", "IRAK1", "IRAK4", "JAK2", "STAT1", "BCL6", "CD40"), drawColBar = FALSE)

# combined IRF pathways
oncoplot(maf = mutations.maf, genes = c("TLR2", "TLR4", "TLR5", "MYD88", "RELA", "NFKB1", "TRAM1", "TICAM1", "TRAF6", "IRAK1", "IRAK4", "JAK2", "STAT1", "BCL6", "CD40", "IRF2BP2", "IRF8", "IRF4", "IRF1"), drawColBar = FALSE)

# IRF family members - for some reason oncoplot does not want to plot genes that have a UTR mutation (but will plot them if that patient has another type of non-UTR mutation) - perhaps try to reduce the maf to just the genes of interest, and add parameter removeNotMutatedGenes to TRUE...
oncoplot(maf = mutations.maf, 
         genes = c("IRF1", "IRF2", "IRF2BP2", "IRF3", "IRF4", "IRF7", "IRF8", "IRF9"), drawColBar = FALSE, 
         colors = c("Frame_Shift_Del"="grey", 
                    "Frame_Shift_Ins"="grey", 
                    "Splice_Site"="orange", 
                    "Translation_Start_Site"="blue",
                    "Nonsense_Mutation"="red", 
                    "Nonstop_Mutation"="red", 
                    "In_Frame_Del"="grey",
                    "In_Frame_Ins"="grey", 
                    "Missense_Mutation"="green3", 
                    "3'UTR"="purple3", 
                    "5'UTR"="dodgerblue3",
                    "Multi_Hit"="black"))

# oncoplot for JAK-STAT pathway
# alternative names: IL2RG ~ IL2gc (gamma chain); PTPN6 ~ SHP1; INPP5D ~ SHIP1; PTPN11 ~ SHP2
oncoplot(maf = mutations.maf, genes = c("IL4R",
                                        "SOCS1",
                                        "STAT6",
                                        "PTPN1",
                                        "IL2RG",
                                        "IL13RA1",
                                        "JAK1", "JAK2", "JAK3",
                                        "CISH",
                                        "PTPN6",
                                        "INPP5D",
                                        "PTPN11"))

# oncoplot for predicted mutually exclusive genes
oncoplot(maf = mutations.maf, genes = c("STAT6", "CISH", "GNA13", "IL4R", "HIST1H1B"), keepGeneOrder = TRUE)

# genes that have significant interactions
sig.genes <- c("STAT6", "CISH", "GNA13", "IL4R", "SOCS1", "CXCR5", "ACTB", "IRF2BP2")

# plot somatic interactions
somaticInteractions(maf = mutations.maf, genes=sig.genes, top = 50, pvalue = c(0.05, 0.01))

# oncoplots for EACH mutually exclusive / co-occurring interaction (most significant):

# mutually exclusive interactions: (30x3)
oncoplot(maf = mutations.maf, genes = c("GNA13", "IL4R"), removeNonMutated = FALSE,
                     drawRowBar = FALSE, drawColBar = FALSE, titleFontSize = 0, fontSize = 15)
oncoplot(maf = mutations.maf, genes = c("CISH", "STAT6"), removeNonMutated = FALSE,
                     drawRowBar = FALSE, drawColBar = FALSE, titleFontSize = 0, fontSize = 15)
oncoplot(maf = mutations.maf, genes = c("CXCR5", "SOCS1"), removeNonMutated = FALSE,
                     drawRowBar = FALSE, drawColBar = FALSE, titleFontSize = 0, fontSize = 15)

# co-occurring interactions:
oncoplot(maf = mutations.maf, genes = c("IRF2BP2", "ACTB"), 
         removeNonMutated = FALSE, 
         drawRowBar = FALSE, 
         drawColBar = FALSE, titleFontSize = 0, fontSize = 15)

laml.titv = titv(maf = mutations.maf, plot = FALSE, useSyn = FALSE)
# plot titv summary
plotTiTv(res = laml.titv)

# Plotting VAF
vafPlot <- plotVaf(maf = mutations.maf, genes = sig.genes, vafCol = 'VAF', 
                   flip = FALSE, top = 50, showN = FALSE)
```

## Lollipop plots

```{r}
library(maftools)
library(gridExtra)

# significant genes only
mutations.maf <- read.maf("/Volumes/shung/projects/PMBCL_exomes/Mutsig/mutsig_input/mutation_file/maftools/paired_and_unpaired.snvs_and_indels.sig_genes_v3.hugo_gene.AF_filtered.gene_filtered.with_AA_Change.maf")

# ALL mutations *including* UTR mutations (and all genes)
mutations.maf <- read.maf("/Volumes/shung/projects/PMBCL_exomes/snv_indel_analysis/maf/mutation_summary-paired_and_unpaired_with_UTRs.maf", vc_nonSyn = c("Frame_Shift_Del", "Frame_Shift_Ins", "Splice_Site", "Translation_Start_Site","Nonsense_Mutation", "Nonstop_Mutation", "In_Frame_Del","In_Frame_Ins", "Missense_Mutation", "3'UTR", "5'UTR"))

mutations.maf <- read.maf("/Volumes/shung/projects/PMBCL_exomes/Mutsig/mutsig_input/mutation_file/maftools/paired_and_unpaired.snvs_and_indels.sig_genes_v3.hugo_gene.AF_filtered.gene_filtered+IRF1_AA_Change.maf")

sig.genes <- c("SOCS1", "ITPKB", "STAT6", "GNA13", "NFKBIE", "PTPN1", "ZNF217", "IL4R", "ACTB", "HIST1H3D", "CD58", "IRF8", "SGPP1", "TNFAIP3", "JUNB", "ZFP36L1", "RGS1", "HIST1H2BC", "HIST1H1C", "JAK1", "MS4A1", "SRSF1", "HIST1H1B", "EZH2", "GDF11", "KLF9", "IL1RAPL2", "SERPINB3", "VMP1", "MYC", "LHFPL3", "TSPAN19", "IRF2BP2", "DDX3X", "CIITA", "IRF4", "CXCR5", "EPHA7", "IL13RA1", "CISH", "SPINK1", "TRAF3", "HIST1H4E", "HIST1H4J", "MAGEC2", "IRX2", "FPGT", "ALX1")

oncoclust.genes <- c("IRF4", "STAT6", "EZH2", "HIST1H3D", "ACTB", "BCL7A")

# create grid of lollipop plots for all significant genes

# for each significant gene, create plot
lolli.plots <- vector('list', length(sig.genes))
for (i in seq_along(sig.genes)) {
  lolli.plots[[i]] <- lollipopPlot(maf = mutations.maf, 
                          gene = sig.genes[i], 
                          AACol = 'Protein_Change', 
                          showMutationRate = TRUE, 
                          domainLabelSize = 2, 
                          defaultYaxis = FALSE
                          )
}
do.call("grid.arrange", c(lolli.plots, ncol=2))

# Structure for proteins: GCSAM, IFNL3, IGLJ2 not found

# Lollipop plots for amino acid changes - issues with retrieving protein position
GNA13.lpop <- lollipopPlot(maf = mutations.maf, 
                          gene = 'GNA13', 
                          AACol = 'Protein_Change', 
                          showMutationRate = TRUE, 
                          domainLabelSize = 3, 
                          defaultYaxis = FALSE, labelPos = c(27, 67, 68)
                          )

# lollipop plots for genes of interest: TBL1XR1, genes in the IRF pathway

```

## Oncodrive to detect cancer driver genes based on positional clustering

Oncodrive is based on the algorithm oncodriveCLUST (originally implemented in python).  The main idea is based on the fact that most of the variants in cancer-causing genes are enriched at few specific loci (aka hot-spots).  This method takes advantage of such positions to identify cancer genes.

```{r}
library(maftools)

# all mutated genes
mutations.maf <- read.maf("/Volumes/shung/projects/PMBCL_exomes/Mutsig/mutsig_input/mutation_file/maftools/paired_and_unpaired.snvs_and_indels.sig_genes.AA_change.maf")

# significant genes only
mutations.maf <- read.maf("/Volumes/shung/projects/PMBCL_exomes/Mutsig/mutsig_input/mutation_file/maftools/paired_and_unpaired.snvs_and_indels.sig_genes_v3.hugo_gene.AF_filtered.gene_filtered.with_AA_Change.maf")

laml.sig = oncodrive(maf = mutations.maf, AACol = 'Protein_Change', minMut = 5, pvalMethod = 'zscore')

# plot the results using plotOncodrive
plotOncodrive(res = laml.sig, fdrCutOff = 0.1, useFraction = TRUE, labelSize = 6)

# the size of the points is proportional to the number of clusters found in the gene
# x-axis: number of mutations (or fraction of mutations) observed in these clusters

```


## Rainfall plot to detect hypermutation

```{r}
library(maftools)
library(gridExtra)

mutations.maf <- read.maf("/Volumes/shung/projects/PMBCL_exomes/Mutsig/mutsig_input/mutation_file/singles_and_pairs/singles_and_pairs.snvs_and_indels.somatic_modified_effects.removed_non_genes.singles_AF_filtered.maf",
                          gisticAllLesionsFile = "/Volumes/shung/projects/PMBCL_exomes/gistic/output/drop_low_coverage-filtered_949-arm_peel-threshold_0.3-fdr_0.05-max_seg_1000-FINAL_FINAL_FINAL/gistic_segments.all_lesions.conf_90.txt", 
                          gisticAmpGenesFile = "/Volumes/shung/projects/PMBCL_exomes/gistic/output/drop_low_coverage-filtered_949-arm_peel-threshold_0.3-fdr_0.05-max_seg_1000-FINAL_FINAL_FINAL/gistic_segments.amp_genes.conf_90.txt", 
                          gisticDelGenesFile = "/Volumes/shung/projects/PMBCL_exomes/gistic/output/drop_low_coverage-filtered_949-arm_peel-threshold_0.3-fdr_0.05-max_seg_1000-FINAL_FINAL_FINAL/gistic_segments.del_genes.conf_90.txt",
                          gisticScoresFile = "/Volumes/shung/projects/PMBCL_exomes/gistic/output/drop_low_coverage-filtered_949-arm_peel-threshold_0.3-fdr_0.05-max_seg_1000-FINAL_FINAL_FINAL/gistic_segments.scores.gistic")

# read in samples that are used as input into gistic (below noisiness threshold of 949)
samples <- read.table("/Volumes/shung/projects/PMBCL_exomes/gistic/input/samples-pass_949.txt")

# rainfall plot for every sample:

#os.plots <- vector('list', length(genes))
#for (i in seq_along(genes)) {
#  os.plots[[i]] <- mafSurvival(maf = mutations.maf, genes = genes[i], clinicalData = clinical.data, addInfo = TRUE, 
#             time = 'Overall.survival..y.', Status = 'CODE_OS', 
#             groupNames = c(paste(genes[i],"-Mutated", sep = ""), "WT"))
#}
#do.call("grid.arrange", c(os.plots, ncol=2))

results <- rainfallPlot(maf = mutations.maf, tsb = c("PA019", "PA043"),
                        detectChangePoints = TRUE, 
                        fontSize = 12, 
                        pointSize = 0.6)

# Filter column in the above results indicate whether the identified segment passes the definition of “Kataegis” which are defined as those genomic segments containing six or more consecutive mutations with an average inter-mutation distance of less than or equal to 1,00 bp 5.

```

## Mutational signature analysis

```{r}
library(maftools)

mutations.maf <- read.maf("/Volumes/shung/projects/PMBCL_exomes/Mutsig/mutsig_input/mutation_file/singles_and_pairs/singles_and_pairs.snvs_and_indels.somatic_modified_effects.removed_non_genes.singles_AF_filtered.maf")

# first, extract adjacent bases to the mutated locus and classify them into 96 subsitution classes
laml.tnm = trinucleotideMatrix(maf = mutations.maf, ref_genome = "~/Documents/references/GRCh37-lite.fa",
                               ignoreChr = "chr23", useSyn = FALSE)

# extractSignatures uses a non-negative matrix factorization to decompose 92 samples x 96 matrix into r signatures.
# By default, the function runs NMF on 6 ranks and chooses the best possible value based on maximum cophenetic-correlation coefficients (can also specify r).  Once decomposed, signatures are compared against known signatures derived from Alexandrov et al., and cosine similarity is calculated to identify the best match.

library(NMF)
laml.sign = extractSignatures(mat = laml.tnm, nTry = 6, plotBestFitRes = FALSE)
#laml.sign = extractSignatures(mat = laml.tnm, nTry = 20, plotBestFitRes = FALSE)

# Using 3 as a best-fit rank based on decreasing cophenetic correlation coefficient.
# Comparing against experimentally validated 30 signatures (See http://cancer.sanger.ac.uk/cosmic/signatures)
# Found Signature_1 most similar to validated Signature_19. Aetiology: Unknown [cosine-similarity: 0.566] 
# Found Signature_2 most similar to validated Signature_3. Aetiology: defects in DNA-DSB repair by HR [cosine-similarity: 0.665] 
#Found Signature_3 most similar to validated Signature_3. Aetiology: defects in DNA-DSB repair by HR [cosine-similarity: 0.556] 

plotSignatures(laml.sign)
plotSignatures(laml.sign, contributions = TRUE)

# look at comparison of similarities of detected signatures against validated signatures
library(corrplot)
corrplot::corrplot(corr = laml.sign$coSineSimMat, 
                   col = RColorBrewer::brewer.pal(n = 9, name = 'Blues'),
                   is.corr = FALSE, 
                   tl.cex = 0.6, 
                   tl.col = 'black', 
                   cl.cex = 0.6)

# alternatively represent in heatmap form:
library(pheatmap)

pheatmap::pheatmap(mat = laml.sign$coSineSimMat, cluster_rows = FALSE, main = "cosine similarity against validated signatures")


```

## Survival analysis

```{r}
library(maftools)
library(gridExtra)

# mutation data for all genes
mutations.maf <- read.maf("/Volumes/shung/projects/PMBCL_exomes/Mutsig/mutsig_input/mutation_file/singles_and_pairs/singles_and_pairs.snvs_and_indels.somatic_modified_effects.removed_non_genes.singles_AF_filtered.maf")

# mutation data filtered for significant genes and rituximab-treated patients only (!)
mutations.maf <- read.maf("/Volumes/shung/projects/PMBCL_exomes/Mutsig/mutsig_input/mutation_file/maftools/paired_and_unpaired.snvs_and_indels.sig_genes_v3.hugo_gene.AF_filtered.gene_filtered.with_VAF.RITU_treated.maf")

clinical.data <- read.table("/Volumes/shung/projects/PMBCL_exomes/data/from_DaveTwa/meta_data_DTwa_updated.txt", sep = "\t", header = TRUE)

# TTP: CODE_TTP and "Time to progression (y)"
# OS: CODE_OS and "Overall survival (y)"
# PFS: CODE_PFS and "Progression free survival (y)"

genes = c("SOCS1", "ITPKB", "STAT6", "GNA13", "NFKBIE", "PTPN1", "ZNF217", "IL4R", "ACTB", "HIST1H3D", "CD58", "IRF8", "SGPP1", "TNFAIP3", "GCSAM", "JUNB", "ZFP36L1", "RGS1", "HIST1H2BC", "HIST1H1C", "JAK1", "MS4A1", "SRSF1", "HIST1H1B", "EZH2", "IFNL3", "GDF11", "KLF9", "IL1RAPL2", "SERPINB3", "VMP1", "MYC", "LHFPL3", "TSPAN19", "IRF2BP2", "DDX3X", "CIITA", "IRF4", "CXCR5", "EPHA7", "IL13RA1", "CISH", "SPINK1", "TRAF3", "HIST1H4E", "HIST1H4J", "MAGEC2", "IRX2", "FPGT")
# omitting IGLJ2A
jak.genes = c("CISH", "IL4R", "IL13RA1", "STAT6", "SOCS1", "PTPN1")

#i<-1
#genes[1] <- irf.genes

# create grid of plots for OS

# for each significant gene, create survival plot
os.plots <- vector('list', length(genes))
for (i in seq_along(genes)) {
  os.plots[[i]] <- mafSurvival(maf = mutations.maf, 
                               genes = genes[i], 
                               clinicalData = clinical.data, 
                               addInfo = TRUE, 
                               time = 'Overall.survival..y.', 
                               Status = 'CODE_OS', 
                               groupNames = c(paste(genes[i],"-Mutated", sep = ""), "WT"))
}
do.call("grid.arrange", c(os.plots, ncol=2))

# create grid of plots for TTP

# for each significant gene, create survival plot
ttp.plots <- vector('list', length(genes))
for (i in seq_along(genes)) {
  ttp.plots[[i]] <- mafSurvival(maf = mutations.maf, 
                                genes = genes[i], 
                                clinicalData = clinical.data, 
                                addInfo = TRUE, 
                                time = 'Time.to.progression..y.', 
                                Status = 'CODE_TTP', 
                                groupNames = c(paste(genes[i],"-Mutated", sep = ""), "WT"))
}
do.call("grid.arrange", c(ttp.plots, ncol=2))

# create grid of plots for PFS

# for each significant gene, create survival plot
pfs.plots <- vector('list', length(genes))
for (i in seq_along(genes)) {
  pfs.plots[[i]] <- mafSurvival(maf = mutations.maf, genes = genes[i], clinicalData = clinical.data, addInfo = TRUE, 
             time = 'Progression.free.survival..y.', Status = 'CODE_PFS', 
             groupNames = c(paste(genes[i],"-Mutated", sep = ""), "WT"))
}
do.call("grid.arrange", c(pfs.plots, ncol=2))


# create separate plots where genes exhibit significant differences in survival between MUT and WT
mafSurvival(maf = mutations.maf, genes = "NFKBIE", clinicalData = clinical.data, 
                                   addInfo = TRUE, time = 'Overall.survival..y.', Status = 'CODE_OS', 
                                   groupNames = c("OS: NFKBIE-Mutated", "WT"), showConfInt = FALSE)
mafSurvival(maf = mutations.maf, genes = "ITPKB", clinicalData = clinical.data, 
                                   addInfo = TRUE, time = 'Time.to.progression..y.', Status = 'CODE_TTP', 
                                   groupNames = c("TTP: ITPKB-Mutated", "WT"), showConfInt = FALSE)
mafSurvival(maf = mutations.maf, genes = "ITPKB", clinicalData = clinical.data, 
                                   addInfo = TRUE, time = 'Progression.free.survival..y.', Status = 'CODE_PFS', 
                                   groupNames = c("PFS: ITPKB-Mutated", "WT"), showConfInt = FALSE)


#hist1h3d.plots <- vector('list', 3)
#hist1h3d.plots[[1]] <- mafSurvival(maf = mutations.maf, genes = "HIST1H3D", clinicalData = clinical.data, 
#                                   addInfo = TRUE, time = 'Overall.survival..y.', Status = 'CODE_OS', 
#                                   groupNames = c("OS: HIST1H3D-Mutated", "WT"), showConfInt = FALSE)
#hist1h3d.plots[[2]] <- mafSurvival(maf = mutations.maf, genes = "HIST1H3D", clinicalData = clinical.data, 
#                                   addInfo = TRUE, time = 'Time.to.progression..y.', Status = 'CODE_TTP', 
#                                   groupNames = c("TTP: HIST1H3D-Mutated", "WT"), showConfInt = FALSE)
#hist1h3d.plots[[3]] <- mafSurvival(maf = mutations.maf, genes = "HIST1H3D", clinicalData = clinical.data, 
#                                   addInfo = TRUE, time = 'Progression.free.survival..y.', Status = 'CODE_PFS',
#                                   groupNames = c("PFS: HIST1H3D-Mutated", "WT"), showConfInt = FALSE)
#do.call("grid.arrange", c(hist1h3d.plots, ncol=3))


```

## GISTIC visualization

```{r}
library(devtools)
install_github("PoisonAlien/maftools")
library(maftools)

all.lesions <- "/Volumes/shung/projects/PMBCL_exomes/gistic/output/drop_low_coverage-filtered_949-arm_peel-threshold_0.3-fdr_0.05-max_seg_1000-FINAL_FINAL_FINAL/gistic_segments.all_lesions.conf_90.txt"
amp.genes <- "/Volumes/shung/projects/PMBCL_exomes/gistic/output/drop_low_coverage-filtered_949-arm_peel-threshold_0.3-fdr_0.05-max_seg_1000-FINAL_FINAL_FINAL/gistic_segments.amp_genes.conf_90.txt"
del.genes <- "/Volumes/shung/projects/PMBCL_exomes/gistic/output/drop_low_coverage-filtered_949-arm_peel-threshold_0.3-fdr_0.05-max_seg_1000-FINAL_FINAL_FINAL/gistic_segments.del_genes.conf_90.txt"
scores.gistic <- "/Volumes/shung/projects/PMBCL_exomes/gistic/output/drop_low_coverage-filtered_949-arm_peel-threshold_0.3-fdr_0.05-max_seg_1000-FINAL_FINAL_FINAL/gistic_segments.scores.gistic"

#source("/Volumes/shung/projects/PMBCL_exomes/gistic/scripts/gisticChromPlot.R")
#source("/Volumes/shung/projects/PMBCL_exomes/gistic/scripts/readSegs.R")

laml.gistic <- maftools::readGistic(gisticAllLesionsFile = all.lesions, 
                          gisticAmpGenesFile = amp.genes, 
                          gisticDelGenesFile = del.genes, 
                          gisticScoresFile = scores.gistic)
# genome plot
gisticChromPlot(laml.gistic)

# gistic oncoplot
col <- c("red", "blue")
names(col) <- c('Amp', 'Del')
gisticOncoPlot(gistic = laml.gistic, colors = col)

```

## Waterfall plot

```{r}
# load the GenVisR package
library("GenVisR")
set.seed(426)

mutations.plot <- mutations.sig
colnames(mutations.plot)[c(1,2)]=c("Hugo_Symbol", "Tumor_Sample_Barcode")

waterfall(mutations.plot, 
          mainRecurCutoff = 0, 
          mainGrid = TRUE,
          plotMutBurden = TRUE, 
          fileType = "MAF", 
          coverageSpace = 90697072,
          plot_proportions = TRUE,
          proportions_type = "TvTi",
          mainDropMut = TRUE,
          rmvSilent = TRUE
          )
#clinData = gender,
#mainXlabel = TRUE,
```

## Analyze IRF mutations

```{r}
library(maftools)
library(gridExtra)
library(dplyr)

# Define the list of genes within the IRF pathway (suggested by Anja + combined with important IRF family members)
irf.genes <- c("TLR2", "TLR4", "TLR5", "MYD88", "RELA", "NFKB1", "TRAM1", "TICAM1", "TRAF6", "IRAK1", "IRAK4", "JAK2", "STAT1", "BCL6", "CD40", "IRF2BP2", "IRF8", "IRF4", "IRF1")

mutations <- read.table("/Volumes/shung/projects/PMBCL_exomes/snv_indel_analysis/mutation_summary-paired_and_unpaired.txt", sep = "\t", header = TRUE)

# extract IRF-familiy mutation data
mutations.irf_family <- unique(rbind(mutations[grep("^IRF", mutations$gene), ]))
mutations.irf_family$gene <- factor(mutations.irf_family$gene)
mutations.irf_family$effect <- factor(mutations.irf_family$effect)
mutations.irf_family.table <- table(mutations.irf_family$gene, mutations.irf_family$effect)
write.table(mutations.irf_family.table, "/Volumes/shung/projects/PMBCL_exomes/pathway_definitions/IRF/irf.mutation.table.txt", sep = "\t", quote = FALSE)

# extract mutation data for the complete IRF pathway
mutations.irf_pathway <- mutations[mutations$gene %in% irf.genes, ]
mutations.irf_pathway$gene <- factor(mutations.irf_pathway$gene)
mutations.irf_pathway$effect <- factor(mutations.irf_pathway$effect)
mutations.irf_pathway.table <- t(table(mutations.irf_pathway$gene, mutations.irf_pathway$effect))
write.table(mutations.irf_pathway, "/Volumes/shung/projects/PMBCL_exomes/pathway_definitions/IRF/mutations.irf_pathway.txt", sep = "\t", quote = FALSE, row.names = FALSE)
write.table(mutations.irf_pathway.table, "/Volumes/shung/projects/PMBCL_exomes/pathway_definitions/IRF/mutations.irf_pathway.table.txt", sep = "\t", quote = FALSE)

# remove genes that are not mutated (not including UTR mutations)
irf.genes.not_mutated <- c("MYD88", "RELA", "TRAM1", "TICAM1", "IRAK1", "CD40")
irf.genes <- irf.genes[ !irf.genes %in% irf.genes.not_mutated ]

# read in MAF files to plot IRF-specific oncoplot and lollipop plots
mutations.maf <- read.maf("/Volumes/shung/projects/PMBCL_exomes/snv_indel_analysis/maf/mutation_summary-paired_and_unpaired_with_UTRs.maf")

lolli.plots <- vector('list', length(irf.genes))
for (i in seq_along(irf.genes)) {
  lolli.plots[[i]] <- lollipopPlot(maf = mutations.maf, 
                          gene = irf.genes[i], 
                          AACol = 'Protein_Change', 
                          showMutationRate = TRUE, 
                          domainLabelSize = 2, 
                          defaultYaxis = FALSE
                          )
}
do.call("grid.arrange", c(lolli.plots, ncol=1))

```

