---
title: "Analysis of outcome correlations for PMBCL exomes"
author: "Stacy Hung"
date: "September 15, 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Survival analysis

The effect of various factors on OS TTP are examined within the context of the PMBCL exomes cohort:
1. MutSigCV genes with a frequency of >= 10%
2. Clinical parameters: Age, Gender, and IPI
3. GISTIC amplifications / deletions

Example questions that we can ask:
- Do patient's age and fitness (IPI) significantly influence the outcome?
- Is a patient's survival affected by mutations in signficant driver genes?

Important terms:
- Event: death or disease recurrence
- "censored" observations (n>=0): patient withdrew from study, lost for follow-up or did not experience event before end of study; cases of non-information

## Process our dataset so it can be used for Survival analysis
```{r}
mutations.sig <- read.table("/Volumes/shung/projects/PMBCL_exomes/Mutsig/mutsig_input/mutation_file/maftools/mutsig_input.FINAL.maf", sep = "\t", header = TRUE)
clinical.df <- read.table("/Volumes/shung/projects/PMBCL_exomes/survival_analysis/clinical_data-extracted_by_Clementine-20180912.txt", sep = "\t", fill = TRUE, header = TRUE)

# binary table for whether or not a gene is mutated in a given patient
mutations.short <- unique(mutations.sig[c("Hugo_Symbol", "Tumor_Sample_Barcode")])
table.gene_by_patient.is_mutated <- table(mutations.short)

write.table(table.gene_by_patient.is_mutated, "/Volumes/shung/projects/PMBCL_exomes/survival_analysis/table.gene_by_patient.is_mutated.txt", sep = "\t", quote = FALSE)

# read back in the data with the two datasets merged
data.df <- read.table("/Volumes/shung/projects/PMBCL_exomes/survival_analysis/clinical_data-with_mutations_in_sig_genes.txt", sep = "\t", fill = TRUE, header = TRUE, as.is = TRUE)

# only look at rituximab-treated patients
data.rit <- subset(data.df, data.df$RITUXIMAB == 1) # 61/94 patients
```

## Load all packages needed for Kaplan-Meier Method and Log-Rank Test

```{r}
library(survival)
library(survminer)
library(dplyr)
library(gridExtra)
```

## Disease-specific surival (DSS)

```{r}
# plot survival curves based on mutated vs. non-mutated patients for each significant gene

# Fit survival data using the Kaplan-Meier method
surv_object <- Surv(time = data.rit$Disease.specific.survival..y.,
                    event = data.rit$CODE_DSS)

# does IPI split nicely into two groups?
hist(data.rit$IPI)

# No- use 0-1, 2-3, and 4-5 as categories
data.rit <- data.rit %>%
  mutate(IPI_group = ifelse(IPI >= 2, "high_intermediate", "low"))

hist(data.rit$IPI_group)

# pass survival data to survival function, and stratify curve based on treatment regimen
#fit1 <- survfit(surv_object ~ EZH2, data = data.rit)
fit1 <- survfit(surv_object ~ IPI_group, data = data.rit)
summary(fit1)

# visualize the corresponding survival curve
ggsurvplot(fit1, data = data.rit, pval = TRUE)

# Fit a Cox proportional hazards model
fit.coxph <- coxph(surv_object ~ IPI + HIST1H3D,
                   data = data.rit)
fit.coxph <- coxph(surv_object ~ IPI + ACTB,
                   data = data.rit)
fit.coxph <- coxph(surv_object ~ IPI + ACTB + HIST1H3D,
                   data = data.rit)
fit.coxph <- coxph(surv_object ~ ACTB + HIST1H3D,
                   data = data.rit)
ggforest(fit.coxph, data = data.rit)


# now for all genes:
genes = c("SOCS1", "ITPKB", "STAT6", "GNA13", "NFKBIE", "PTPN1", "ZNF217", "IL4R", "ACTB", "HIST1H3D", "CD58", "IRF8", "SGPP1", "TNFAIP3", "GCSAM", "JUNB", "ZFP36L1", "RGS1", "HIST1H2BC", "HIST1H1C", "JAK1", "MS4A1", "SRSF1", "HIST1H1B", "EZH2", "IFNL3", "GDF11", "KLF9", "IL1RAPL2", "SERPINB3", "VMP1", "MYC", "LHFPL3", "TSPAN19", "IRF2BP2", "DDX3X", "CIITA", "IRF4", "CXCR5", "EPHA7", "IL13RA1", "CISH", "SPINK1", "TRAF3", "HIST1H4E", "HIST1H4J", "MAGEC2", "IRX2", "FPGT")

#km.dss.plots <- vector('list', length(genes))
fits <- vector('list', length(genes)) # list of survfit objects
legend <- vector('list', length(genes))

for (i in seq_along(genes)) {
  # approach 1
  modform <- as.formula(paste("surv_object", genes[i], sep = " ~ "))
                               substitute(survfit(modform, data = data.rit), list(modform = modform))
  fits[[i]] <- eval(substitute(survfit(modform, data = data.rit), list(modform = modform)))
  legend[i] <- genes[i]
  
  # original approach
  #fit <- survfit(surv_object ~ factor(genes[i]), data = data.rit)
  #summary(surv_object.dss[i])
  # visualize the corresponding survival curve
  #km.dss.plots[[i]] <- ggsurvplot(fit, data = data.rit, pval = TRUE)
}
ggsurvplot(fits[i], data.rit, pval = TRUE)
ggsurvplot_list(fits, data.rit, legend.title = legend)

# plot genes that have a signficant difference in survival
NFKBIE_INDEX <- 5
IL4R_INDEX <- 8
ACTB_INDEX <- 9
HIST1H3D_INDEX <- 10

ggsurvplot(fits[NFKBIE_INDEX], data.rit, pval = TRUE, conf.int = TRUE, pval.method = TRUE, risk.table = TRUE, cumevents = TRUE, tables.height = 0.2)

ggsurvplot(fits[ACTB_INDEX], data.rit, pval = TRUE, conf.int = TRUE, pval.method = TRUE, risk.table = TRUE, cumevents = TRUE, tables.height = 0.2)

ggsurvplot(fits[HIST1H3D_INDEX], data.rit, pval = TRUE, conf.int = TRUE, pval.method = TRUE, risk.table = TRUE, cumevents = TRUE, tables.height = 0.2)

ggsurvplot(fits[IL4R_INDEX], data.rit, pval = TRUE, conf.int = TRUE, pval.method = TRUE, risk.table = TRUE, cumevents = TRUE, tables.height = 0.2)
```

## Time-to-progression TTP

```{r}
# plot survival curves based on mutated vs. non-mutated patients for each significant gene

# Fit survival data using the Kaplan-Meier method
surv_object <- Surv(time = data.rit$Time.to.progression..y.,
                    event = data.rit$CODE_TTP)

# pass survival data to survival function, and stratify curve based on treatment regimen
fit1 <- survfit(surv_object ~ EZH2, data = data.rit)
summary(fit1)

# visualize the corresponding survival curve
ggsurvplot(fit1, data = data.rit, pval = TRUE)

# now for all genes:
genes = c("SOCS1", "ITPKB", "STAT6", "GNA13", "NFKBIE", "PTPN1", "ZNF217", "IL4R", "ACTB", "HIST1H3D", "CD58", "IRF8", "SGPP1", "TNFAIP3", "GCSAM", "JUNB", "ZFP36L1", "RGS1", "HIST1H2BC", "HIST1H1C", "JAK1", "MS4A1", "SRSF1", "HIST1H1B", "EZH2", "IFNL3", "GDF11", "KLF9", "IL1RAPL2", "SERPINB3", "VMP1", "MYC", "LHFPL3", "TSPAN19", "IRF2BP2", "DDX3X", "CIITA", "IRF4", "CXCR5", "EPHA7", "IL13RA1", "CISH", "SPINK1", "TRAF3", "HIST1H4E", "HIST1H4J", "MAGEC2", "IRX2", "FPGT")

#km.dss.plots <- vector('list', length(genes))
fits <- vector('list', length(genes)) # list of survfit objects
legend <- vector('list', length(genes))

for (i in seq_along(genes)) {
  # approach 1
  modform <- as.formula(paste("surv_object", genes[i], sep = " ~ "))
                               substitute(survfit(modform, data = data.rit), list(modform = modform))
  fits[[i]] <- eval(substitute(survfit(modform, data = data.rit), list(modform = modform)))
  legend[i] <- genes[i]
  
  # original approach
  #fit <- survfit(surv_object ~ factor(genes[i]), data = data.rit)
  #summary(surv_object.dss[i])
  # visualize the corresponding survival curve
  #km.dss.plots[[i]] <- ggsurvplot(fit, data = data.rit, pval = TRUE)
}
ggsurvplot(fits[i], data.rit, pval = TRUE)
ggsurvplot_list(fits, data.rit, legend.title = legend)

NFKBIE_INDEX <- 5
IL4R_INDEX <- 8
ACTB_INDEX <- 9
HIST1H3D_INDEX <- 10
ITPKB_INDEX <- 2

# no longer significant in TTP (but were found to be signficant in DSS)
ggsurvplot(fits[NFKBIE_INDEX], data.rit, pval = TRUE)
ggsurvplot(fits[ACTB_INDEX], data.rit, pval = TRUE)
ggsurvplot(fits[HIST1H3D_INDEX], data.rit, pval = TRUE)

ggsurvplot(fits[ITPKB_INDEX], data.rit, pval = TRUE, conf.int = TRUE, pval.method = TRUE, risk.table = TRUE, cumevents = TRUE, tables.height = 0.2)

ggsurvplot(fits[IL4R_INDEX], data.rit, pval = TRUE, conf.int = TRUE, pval.method = TRUE, risk.table = TRUE, cumevents = TRUE, tables.height = 0.2)

```

## Cox Proportional Hazards Models

```{r}

```

## Implementation of Survival analysis in R (pilot using real data for ovarian cancer)

columns of interest:

futime : time patients were tracked until they either died or were lost to follow-up (*survival times - response var)
fustat : whether patients were censored or not
age 
rx : treatment group assigned - patient recieves 1 or 2 therapy regimens
resid.ds : presence of residual disease
ecog.ps : performance status (ECOG)

```{r}
library(survival)
library(survminer)
library(dplyr)

# Import dataset
data(ovarian)
glimpse(ovarian)

# if we want to break up cohort into age categories, we need to know the distribution:
# first, dichotomize age and change data labels
ovarian$rx <- factor(ovarian$rx,
                     levels = c("1", "2"),
                     labels = c("A", "B"))

ovarian$resid.ds <- factor(ovarian$resid.ds,
                           levels = c("1", "2"),
                           labels = c("no", "yes"))

ovarian$ecog.ps <- factor(ovarian$ecog.ps,
                          levels = c("1", "2"),
                          labels = c("good", "bad"))

hist(ovarian$age)

ovarian <- ovarian %>% mutate(age_group = ifelse(age >=50, "old", "young"))
ovarian$age_group <- factor(ovarian$age_group)

# Fit survival data using the Kaplan-Meier method
surv_object <- Surv(time = ovarian$futime,
                    event = ovarian$fustat)

# pass survival data to survival function, and stratify curve based on treatment regimen
fit1 <- survfit(surv_object ~ rx, data = ovarian)
summary(fit1)

# visualize the corresponding survival curve
ggsurvplot(fit1, data = ovarian, pval = TRUE)

# by convention, vertical lines indicate censored data, their corresponding x values is the time at which the censoring occurred

# survival did not differ significantly by treatment, but what about other variables?

# examine predictive value of residual disease status
fit2 <- survfit(surv_object ~ resid.ds, data = ovarian)
ggsurvplot(fit2, data = ovarian, pval = TRUE)

# Is there a systematic way to look at the different covariates? Cox proportional hazards models
# Briefly, a HR > 1 indicates an increased risk of death if a specific condition is met by a patient; HR < 1, on the other hand, would indicate a decreased risk.

# Fit a Cox proportional hazards model
fit.coxph <- coxph(surv_object ~ rx + resid.ds + age_group + ecog.ps,
                   data = ovarian)
ggforest(fit.coxph, data = ovarian)

```